Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-02-05 article/01_Introduction.Rmd
even more restructuring of the introduction -- some aspects where moved to the conclusion

Burial traditions are a ritual behaviour that is transmitted between generations. They are therefore suitable for study using methods of *Cultural transmission theory*, a special aspect of *Cultural evolution* concerned with the spread of ideas in human populations [@creanzaCulturalEvolutionaryTheory2017]. This theoretical framework permits the comparison of behaviour across spatial, temporal and cultural boundaries. *Evolutionary archaeology*, the branch of archaeology that explicitly adopted the terminology of Cultural evolution theory, established an important distinction of cultural behaviour in *function* and *style* [@dunnell1978style] which is very relevant for burial traditions: Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their diffusion can be expected to increase over time if we assume reasonable human actors. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. They serve archaeologists as a proxy for precise relative chronology. Burial traditions can be descriped as a special form of *stylistic traits*. Special because of their strong dependence on religious and spiritual beliefs and their connection with the emotionally stirring topic of death, but still subject to seemingly random fashion changes.

@@

In the following sections I will (1) *explain how radiocarbon data with relevant meta information can be transformed to macro regional time series of burial traditions*, (2) *explore how these time series compare to traditional archaeological narratives* and (3) *construct and contextualise cultural distances between macro regions over time in relation to spatial distance*. For the latter a simulation based on the @neimanStylisticVariationEvolutionary1995 template was employed.

Cultural evolution theory is at least as old as theories of biological evolution and underwent a long and versatile research history. Its beginnings are in *Evolutionism* and *Social darwinism* in the 19th and early 20th century and until the nineteen sixties and seventies it matured in conflict with *Cultural relativism* into *Neoevolutionism* [@petermannGeschichteEthnologie2004]. In the late seventies three major theoretical movements emerged: *Evolutionary psychology*, *Human behavioural ecology* and *Dual inheritance theory* [@smithThreeStylesEvolutionary2000]. The latter -- heavily popularised by Richard Dawkins's *Memetics* [@dawkinsSelfishGene1976] -- establishes the notion of culture acting as secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007]: Just like in biological evolution, also *Social learning* might be prone to selection and mutation. Under this assumption methods of *Population genetics* could be employed within cultural anthropology to analyse the spread and development of ideas -- of course with the much faster pace of cultural exchange in mind. A lot of theoretical and quantitative research followed this line of thought: e.g. @cavalli-sforzaCulturalTransmissionEvolution1981, @boydCultureEvolutionaryProcess1985, @teltserEvolutionaryArchaeologyMethodological1995 or @obrienEvolutionaryArchaeologyTheory1996.

Archaeology has a unique perspective on Cultural evolution as it is concerned with human behaviour over big areas and long time scales [@garveyCurrentPotentialRoles2018]. It focuses on an abundant and highly diagnostic proxy of cultural expression: material remains. These include individual lost or discarded artefacts, the ruins of past architecture and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can, therefore, be explained as a direct result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. Questions of Cultural Transmission form the core of archaeology.

>>>>

Burial traditions are a ritual behaviour that is transmitted between generations. They are therefore suitable for study using methods of *Cultural transmission theory*, a special aspect of *Cultural evolution theory* concerned with the spread of ideas in human populations [@creanzaCulturalEvolutionaryTheory2017]. This theoretical framework permits the comparison of behaviour across spatial, temporal and cultural boundaries, as it assumes *Social learning* to produce adaptive behaviour through transmission, selection and mutation -- a secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007; @lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002].

*Evolutionary archaeology*, the branch of archaeology that explicitly adopted the terminology of Cultural evolution, established an important distinction of cultural behaviour in *function* and *style* [@dunnell1978style] which is very relevant for burial traditions: Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their prevalence generally happens to increase over time. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. Burial traditions can be descriped as a special form of these *stylistic traits*. Special because of their strong dependence on religious and spiritual beliefs and their connection with the emotionally stirring topic of death, but still subject to sometimes rapid and seemingly random fashion changes.

@@

In the following sections I will (1) *explain how radiocarbon data with relevant meta information can be transformed to macro regional time series of burial traditions*, (2) *explore how these time series compare to traditional archaeological narratives* and (3) *construct and contextualise cultural distances between macro regions over time in relation to spatial distance*.

### 2019-02-05 article/05_Conclusion.Rmd
even more restructuring of the introduction -- some aspects where moved to the conclusion

2. This paper is built upon Cultural Evolution theory (see section \ref{introduction}). It not just relies on terminology and methodology developed in this field, the process of abstraction necessary to go from radiocarbon dates to cultural distance was also heavily influenced by Evolutionary archaeology. This theoretical framework might be a source of inspiration and methods for other projects. The adoption and adaption of statistical tools and models from population genetics into anthropology was and is a unique characteristic of Cultural Evolution theory [@creanzaCulturalEvolutionaryTheory2017].

>>>>

2. This paper is built upon Cultural evolution theory (see section \ref{introduction}). It not just relies on terminology and methodology developed in this field, the process of abstraction necessary to go from radiocarbon dates to cultural distance was also heavily influenced by Evolutionary archaeology. This theoretical framework might be a source of inspiration and methods for other projects. The adoption and adaption of statistical tools and models from population genetics into anthropology was and is a unique characteristic of Cultural Evolution theory [@cavalli-sforzaCulturalTransmissionEvolution1981; @boydCultureEvolutionaryProcess1985; @teltserEvolutionaryArchaeologyMethodological1995; @obrienEvolutionaryArchaeologyTheory1996]. On the other hand, archaeology is explicitly concerned with human behaviour over big areas and long time scales and can offer a unique perspective to other anthropological fields [@garveyCurrentPotentialRoles2018].
